# Vectors algebra

## Libraries

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(magrittr)
```

## Vector Mathematics in R

R vectors support basic arithmetic operations that correspond to the same operations on geometric vectors. For example:

```{r, prompt=TRUE, collapse=TRUE, results='markup'}
x <- 1:15
y <- 10:24
x
y
x + y # vector addition
x - y # vector subtraction
x * 3 # multiplication by a scalar
```



R also has an operator for the dot product, denoted `%*%`.
This operator also designates matrix multiplication, which we will
discuss in the next chapter. By default this operator returns an object of the R
matrix class. If you want a scalar (or the R equivalent of a scalar,
i.e. a vector of length 1) you need to use the `drop()` function.

```{r, prompt=TRUE, collapse=TRUE, results='markup'}
z <- x %*% x
class(z)      # note use of class() function
z
drop(z)
```


In lecture we saw that many useful geometric properties of vectors could be expressed in the form of dot products. Let's start with some two-dimensional vectors where the geometry is  easy to visualize:

```{r, prompt=TRUE, collapse=TRUE, results='markup'}
a <- c(2, 0)   # the point (2,0)
b <- c(1, 3)   # the point (1,3)
```

To draw our vectors using ggplot, we'll need to create a data frame with columns representing the x,y coordinates of the end-points of our vectors: 

```{r}
df <- data.frame(x.end = c(a[1], b[1]),
                 y.end = c(a[2], b[2]),
                 label = c('a', 'b'))

ggplot(df) + 
  geom_segment(aes(x=0, y = 0, xend = x.end, yend = y.end, color=label), 
               arrow = arrow()) +
  labs(x = "x-coordinate", y = "y-coordinate") + 
  coord_fixed(ratio = 1) + # insures x and y axis scale are same
  theme_bw()
```



Let's see what the dot product can tell us about these vectors. First recall that we can calculate the length of a vector as the square-root of the dot product of the vector with itself ($\vert\vec{a}\vert^2  =  \vec{a} \cdot \vec{a}$)

```{r, prompt=TRUE, collapse=TRUE, results='markup'}
len.a <- drop(sqrt(a %*% a))
len.a
len.b <- drop(sqrt(b %*% b))
len.b
```

How about the angle between $a$ and $b$?  First we can use the dot product and the previously calculated lengths to calculate the cosine of the angle between the vectors:

```{r, prompt=TRUE, collapse=TRUE, results='markup'}
cos.ab <- (a %*% b)/(len.a * len.b)
cos.ab
```

To go from the cosine of the angle to the angle (in radians) we need the arc-cosine function, `acos()`:

```{r, prompt=TRUE, collapse=TRUE, results='markup'}
acos(cos.ab)  # given angle in radians
```



## Simple statistics in vector form

Now let's turn our attention to seeing how to calculate a variety of simple statistics such as the mean, variance, etc. in terms of vector operations. 

To illustrate these oeprations we'll use the I. setosa data from the `iris` examplar data set.

```{r}
setosa <- filter(iris, Species == "setosa")
```

### Mean

First let's calculate the mean for the Sepal.Length variable. Referring back to the slides for today's lecture, we see we can calculate the mean as:
$$
\bar{x} = \frac{\vec{1} \cdot \vec{x}}{\vec{1} \cdot \vec{1}}
$$

Applying this formula in R:

```{r, prompt=TRUE, collapse=TRUE, results='markup'}
sepal.length <- setosa$Sepal.Length
ones <- rep(1, length(sepal.length)) # 1-vector of length n
mean.sepal.length <- (ones %*% sepal.length)/(ones %*% ones)
mean.sepal.length %<>% drop  # use drop to convert back to scalar
mean.sepal.length
```

Let's compare our calculation against the built-in mean function:

```{r, prompt=TRUE, collapse=TRUE, results='markup'}
mean(sepal.length)
```

### Mean centering

Mean centering a vector, means subtracting the mean from each element of that vector:

$$
\vec{x}_c = \vec{x} - \bar{x}\vec{1}
$$

Now let's create a mean centered vector from sepal.length, which we'll refer to as the vector of deviates about the mean:

```{r, prompt=TRUE, collapse=TRUE, results='markup'}
sepal.length.deviates <- sepal.length - mean.sepal.length  
```

Note that we didn't have to explicitly multiply the a one vector by the mean, as R will automatically make the lengths of the `sepal.length` (a vector of length 150) and `mean.sepal.length` (a vector of length 1) match by vector recycling.

### Variance and standard deviation

Using the vector of deviates we can easily calculate the variance and standard deviation of a variable.

The variance of a variable, in vector algebraic terms, is:

$$
S_x^2 = \frac{\vec{x}_c \cdot \vec{x}_c}{n-1}
$$

The standard deviation is simply the square root of the variance

$$
S_x = \sqrt{S_x^2} 
$$

These calculations for the Sepal.Length variable:

```{r, prompt=TRUE, collapse=TRUE, results='markup'}
n <- length(sepal.length.deviates)
var.sepal.length <- (sepal.length.deviates %*% sepal.length.deviates)/(n-1) 
var.sepal.length
sd.sepal.length <- sqrt(var.sepal.length)
sd.sepal.length
```

Again, we can compare our calculations to the built-in `var()` and `sd()` functions:

```{r, prompt=TRUE, collapse=TRUE, results='markup'}
var(sepal.length)
sd(sepal.length)
```

### Covariance and correlation

Now let's consider the common measures of bivariate association, covariance and correlation.

Covariance is:

$$
S_{XY} = \frac{\vec{x} \cdot \vec{y}}{n-1}
$$

Correlation is:

$$
r_{XY} = \frac{\vec{x} \cdot \vec{y}}{|\vec{x}||\vec{y}|} = \frac{S_{XY}}{S_x S_Y}
$$


We'll examine the relationship between sepal length and width:

```{r}
sepal.width <- setosa$Sepal.Width
mean.sepal.width <-  (ones %*% sepal.width)/(ones %*% ones) 
sepal.width.deviates <- sepal.width - mean.sepal.width
var.sepal.width <- drop((sepal.width.deviates %*% sepal.width.deviates)/(n-1))
sd.sepal.width <- sqrt(var.sepal.width)
```

With the vector of sepal width deviates in hand we can now calculate covariances:

```{r, prompt=TRUE, collapse=TRUE, results='markup'}
cov.swidth.slength <- (sepal.length.deviates %*% sepal.width.deviates)/(n-1)
cov.swidth.slength
cov(sepal.length, sepal.width) # and compare to built-in covariance
```

And correlations:

```{r, prompt=TRUE, collapse=TRUE, results='markup'}
len.sepal.length <- sqrt(sepal.length.deviates %*% sepal.length.deviates)
len.sepal.width <- sqrt(sepal.width.deviates %*% sepal.width.deviates) 

corr.swidth.slength <- 
  (sepal.length.deviates %*% sepal.width.deviates) / (len.sepal.length * len.sepal.width)
corr.swidth.slength
cor(sepal.length, sepal.width) # and compare to built-in correlation
```

Alternately, we could have calculated the correlation more simply as follows:

```{r, prompt=TRUE, collapse=TRUE, results='markup'}
cov.swidth.slength/(sd.sepal.length * sd.sepal.width)
```
