
# Matrices in R

In R matrices are two-dimensional collections of elements all of which
have the same mode or type. This is different than a data frame in which
the columns of the frame can hold elements of different type (but all of
the same length), or from a list which can hold objects of arbitrary
type and length. Matrices are more efficient for carrying out most
numerical operations, so if you're working with a very large data set
that is amenable to representation by a matrix you should consider using
this data structure.

```{r, warning=FALSE, message=FALSE}
library(tidyverse)
```


## Creating matrices in R

There are a number of different ways to create matrices in R. For
creating small matrices at the command line you can use the
`matrix()` function.


```{r, prompt=TRUE, collapse=TRUE, results='markup'}
x <- matrix(1:5) # creates a column vector
x
X <- matrix(1:12, nrow=4) # creates a matrix
X
dim(X) # give the shape of the matrix
```

`matrix()` takes a data vector as input and the shape of the
matrix to be created is specified by using the `nrow` and
`ncol` arguments. If the number of elements in the input data
vector is less than `nrows` $\times$ `ncols` the
elements will be 'recycled' as discussed in previous chapters. Without
any shape arguments the `matrix()` function will create a
column vector as shown above. By default the `matrix()`
function fills in the matrix in a column-wise fashion. To fill in the
matrix in a row-wise fashion use the argument `byrow=T`.

If you have a pre-existing data set in a list or data frame you can use
the `as.matrix()` function to convert it to a matrix.

```{r, prompt=TRUE, collapse=TRUE, results='markup'}
iris.mtx <- as.matrix(iris)
head(iris.mtx) # NOTE: the elements were all converted to character
```

Since all elements of an R matrix must be of the same type, when we passed the iris data frame to `as.matrix()`, everything was converted to a character due to the presence of the Species column in the data frame.

```{r, prompt=TRUE, collapse=TRUE, results='markup'}
# This is probably more along the lines of what you want
iris.mtx <- iris %>% dplyr::select(-Species) %>% as.matrix
head(iris.mtx)
```

You can use the various indexing operations to get particular rows,
columns, or elements. Here are some examples:

```{r, prompt=TRUE, collapse=TRUE, results='markup'}
X <- matrix(1:12, nrow=4)
X
X[1,] # get the first row
X[,1] # get the first column
X[1:2,] # get the first two rows
X[,2:3] # get the second and third columns
Y <- matrix(1:12, byrow=T, nrow=4)
Y
Y[4] # see explanation below
Y[5]
dim(Y) <- c(2,6)  # reshape Y
Y
Y[5]
```

The example above where we create a matrix `Y` is meant to
show that matrices are stored internally in a column wise fashion (think
of the columns stacked one atop the other), regardless of whether we use
the `byrow=T` argument. Therefore using single indices returns
the elements with respect to this arrangement. Note also the use of
assignment operator in conjuction with the `dim()` function to
reshape the matrix. Despite the reshaping, the internal representation
in memory hasn't changed so `Y[5]` still gives the same
element.

You can use the `diag()` function to get the diagonal of a
matrix or to create a diagonal matrix as show below:

```{r, prompt=TRUE, collapse=TRUE, results='markup'}
Z <- matrix(rnorm(16), ncol=4)
Z
diag(Z)
diag(5) # create the 5 x 5 identity matrix
s <- sqrt(10:13)
diag(s)
```


## Matrix arithmetic operations in R

The standard mathematical operations of addition and subtraction and
scalar multiplication work element-wise for matrices in the same way as
they did for vectors. Matrix multiplication uses the operator
`%*%` which you saw last week for the dot product. To get the
transpose of a matrix use the function `t()`. 

```{r, prompt=TRUE, collapse=TRUE, results='markup'}
A <- matrix(1:12, nrow=4)
A

B <- matrix(rnorm(12), nrow=4)
B

t(A)  # transpose
A + B  # matrix addition
A - B  # matrix subtraction
5 * A  # multiplication by a scalar
```

When applying matrix multiplication, the dimensions of the matrices involved must be conformable. For example, you can't do this:

```
A %*% B  # do you understand why this generates an error?
```

But this works:
```{r, prompt=TRUE, collapse=TRUE, results='markup'}
A %*% t(B)
```



## Descriptive statistics as matrix functions

Assume you have a data set represented as a $n \times p$ matrix, $X$, with
observations in rows and variables in columns. Below I give formulae for
calculating some descriptive statistics as matrix functions.

### Mean vector and matrix

You can calculate a row vector of means, $\mathbf{m}$, as: 
$$
\mathbf{m} = \frac{1}{n} \mathbf{1}^T  X
$$ 
where $1$ is a $n \times 1$ vector of ones.

A $n \times p$ matrix $M$ where each column is filled with the mean
value for that column is: 
$$
M = \mathbf{1}\mathbf{m}
$$

### Deviation matrix

To re-express each value as the deviation from the variable means
(i.e.~each columns is a mean centered vector) we calculate a deviation
matrix: 
$$
D = X - M
$$

### Covariance matrix

The $p \times p$ covariance matrix can be expressed as a matrix product of the deviation matrix:
$$
S = \frac{1}{n-1} D^T D
$$

### Correlation matrix

The correlation matrix, $R$, can be calculated from the covariance
matrix by: 
$$
R = V S V
$$

where $V$ is a $p \times p$ diagonal matrix where
$V_{ii} = 1/\sqrt{S_{ii}}$.


## Matrix Inverse

The function `solve()` can be used to find matrix inverses in R.

```{r}
A <- matrix(1:4, nrow=2)
A

Ainv <- solve(A)
Ainv

A %*% Ainv   # should give identity matrix
Ainv %*% A   # should also result in identity matrix
```

Keep in mind that not all square matrices are invertible:

```{r, error=TRUE}
C <- matrix(1:16, nrow=4)
C

Cinv <- solve(C)
```
