
knitr::opts_chunk$set(echo = TRUE, prompt = FALSE, eval = TRUE, 
                      warning = FALSE, comment=NA, cache = FALSE,
                      fig.width = 6, fig.height = 4, results = "markup")

# Data exploration: First steps

In this chapter we're going to make our first forays into exploratory data analysis. The primary goal of  exploratory data analysis to gain a good understanding of the data of interest, such as distributions, trends, and outliers. Usually this involves data visualization followed by simple summary statistics to summarize features like location, spread, covariation among variables, etc.


## Load the `tidyverse` package

We'll be using functions defined in three packages: 1) `readr` for reading "rectangular data" (e.g. spreadsheets); 2) `dplyr` for data filtering and summarizing; and  3) `ggplot2` for creating data visualizations.  All three of these packages are part of the the "tidyverse" meta-package, so we can load them simultaneously by loading tidyverse:

```{r}
library(tidyverse)
```


## Reading data with the `readr` package

The [`readr`](http://readr.tidyverse.org) package defines a number of functions for reading data tables from common file formats like Comma-Separated-Value (CSV) and Tab-Separated-Value (TSV) files.

For the purposes of illustration we'll use a data set I'll refer to as the "possums data set"  The possums data set includes information from a study of mountain brushtail possums (*Trichosurus caninus*; Lindenmayer DB et al. 1995, Australian Journal of Zoology 43, 449-458.)  The investigators recorded variables about individual possum's sex, age, where they were collected, and a range of morphological measurements. For example the variable `skullw` is the width of the skull, `taill` is tail length, etc. This data is available at https://tinyurl.com/possums-csv.

We'll use the `read_csv()` function to load a CSV formatted file with this data:

```{r, resulds="markup"}
possums <- read_csv("https://tinyurl.com/possums-csv")
```

Notice that when you used `read_csv()` the function printed information about how it "parsed" the data (i.e. the types it assigned to each of the columns).

Notice too that we read the CSV file directly from a remote file via a URL.  If instead, you wanted to load a local file on your computer you would specify the "path" -- i.e. the location on your hard drive where you stored the file.  For example, here is how I would load the same file if it was stored in the Downloads directory on my Mac laptop:

```{r, eval = FALSE}
# this assumes you're running a local version of R (i.e. on your laptop)
possums <- read_csv("/Users/pmagwene/Downloads/possums.csv")
```


## Structure of the possums data

Let's take a moment to examine the structure of possums data set.  

First, how big is it? The `dim` function will tell us the dimensions (# of rows and columns) of the data table we loaded:

```{r}
dim(possums)
```
What are the names of the columns of the table?

```{r}
names(possums)
```

What is the type of the object?

```{r}
typeof(possums)
```

You may recall that both lists and data frames have have the "list" data type.  The `class()` function distinguishes lists from data frames:

```{r}
class(possums)
```

Here we see that `possums` object actually has three classes -- `tbl_df`, `tbl`, and `data.frame`.  For most practical purposes we can treat objects returned by `read_csv()` and the other `reader` functions as data frames.



## Simple tables

Let's use the `count` function to count the number of male and female possums and the number collected from each of the populations of interest in this study (Victoria and "other"):

```{r}
count(possums, sex)
count(possums, Pop)
```

We can break down the counts by population and sex combinined by specifying both variables:

```{r}
count(possums, Pop, sex)
```
How do the results differ if  instead you write `count(possums, sex, Pop)`?

Finally, let's get counts for the different age groups in the study:

```{r}
count(possums, age)
```

Notice that besides the nine age groups, there is a grouping for "NA". "NA" means "not available" and is the standard designation for missing values.  This table tells us there are two possums for which age estimates were not available.


## Figures using `ggplot2`

Throughout this course we'll use a package called "ggplot2" to generate figures.  The ggplot2 package is part of the tidyverse and is automatically loaded when we load the tidyverse library.  In the code below, we'll demonstrate how to use ggplot by example.  In a later lecture we'll go into greater detail about how ggplot is structured and the broader conceptual framework that underpins its design.

### Bar plots

First let's create some simple bar plots as alternate representations of the counts:

```{r}
ggplot(possums) + geom_bar(aes(x = sex))
```

```{r}
ggplot(possums) + geom_bar(aes(x = Pop))
```

```{r}
ggplot(possums) + geom_bar(aes(x = Pop, fill = sex))
```


### Histograms

A histogram is a special type of bar plot, typically used with continuous data.  In a histogram, we divide the range of the data into bins of a given size, and use vertical bars to depict the frequency (count) of observations that fall into each bin.  This gives a good sense of the intervals in which most of the observations are found.

Here is a histogram for the skull width data:

```{r}
ggplot(possums) + geom_histogram(aes(x = skullw))
```

Notice the warning message that was generated, about the default number of bins that were used.  ggplot is alerting us that that we may want to consider regenerating the plot with a different number of bins.  Let's try a smaller number of bins:

```{r}
ggplot(possums) + geom_histogram(aes(x = skullw), bins = 20)
```

Try creating histograms for `taill` and `totlngth`.

### Scatter plots

A scatter plot is typically used to represent two numerical variables simultaneously. Each point in a scatter plot is an individual in the data set, and the location of the points represent the measured values of the variables of interest on that individual.

```{r}
ggplot(possums) + geom_point(aes(x = totlngth, y = taill))
```

We can add information about categorical variables to our scatter plot by using color or shape to depict different classes


```{r}
ggplot(possums) + 
  geom_point(aes(x = totlngth, y = taill, color = Pop))
```


We can represent more than two categorical variables in our scatter plot by using both shape and color.  We'll also change the size and transparency of the plotted points (via the `alpha` argument).

```{r}
ggplot(possums) + 
  geom_point(aes(x = totlngth, y = taill, color = Pop, shape = sex),
             size = 2, alpha = 0.75)
```

Explore some of the other bivariate relationships in the possums data by creating additional scatter plots.  Given 9 numerical variables, how many distinct pairwise scatter plots could you create?

